---
title: "Lab 2"
author: "Amritpal Sandhu, Collaborator: Elise Gonzales"
date: "2023-01-24"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library("tidymodels")
library("tidyverse")
library("dplyr")
library("janitor")
library("corrplot")
dat <- read_csv(file = "https://raw.githubusercontent.com/MaRo406/eds-232-machine-learning/main/data/pumpkin-data.csv")
```

```{r data, include=FALSE}
glimpse(dat)
```

```{r, include=FALSE}
# Clean names to the snake_case convention

pumpkins <- dat %>% clean_names(case = "snake")

# Return column names

pumpkins %>% names()
```


```{r, include=FALSE}
pumpkins <- pumpkins %>% select(variety, city_name, package, low_price, high_price, date)

## Print data set

pumpkins %>% slice_head(n = 5)
```

```{r, include=FALSE}
## Load lubridate

library(lubridate)

# Extract the month and day from the dates and add as new columns
pumpkins <- pumpkins %>%
  mutate(date = mdy(date),  
         day = yday(date),
         month = month(date))
pumpkins %>% 
  select(-day)

## View the first few rows

pumpkins %>% slice_head(n = 7)
```


```{r, include=FALSE}
# Create a new column price
pumpkins <- pumpkins %>% 
  mutate(price = (low_price+ high_price)/2)
```


```{r, echo=TRUE, include=FALSE}
ggplot(data = pumpkins, aes(x = day, y = price)) + geom_point(color = "blue", 
             alpha = .2, 
             size = 3) +
  labs(x = "Day", y = "Price", title = "Pumpkin Sales Throughout the Year")
```

```{r, include=FALSE}
# Verify the distinct observations in Package column
pumpkins %>% 
  distinct(package)
```


```{r, echo=TRUE, include=FALSE}
# Retain only pumpkins with "bushel" in the package column
new_pumpkins <- pumpkins %>% filter(str_detect(package, "bushel"))

# Get the dimensions of the new data
dim(new_pumpkins)

# View a few rows of the new data
new_pumpkins %>% 
  slice_head(n = 10)
```

```{r, include=FALSE}
# Convert the price if the Package contains fractional bushel values
new_pumpkins <- new_pumpkins %>% 
  mutate(price = case_when(
    str_detect(package, "1 1/9") ~ price/(1.1),
    str_detect(package, "1/2") ~ price*2,
    TRUE ~ price))

# View the first few rows of the data
new_pumpkins %>% 
  slice_head(n = 30)
```


```{r, include=FALSE}
# Set theme
theme_set(theme_light())

# Make a scatter plot of month and price
new_pumpkins %>% 
  ggplot(mapping = aes(x = day, y = price)) +
  geom_point(size = 1.6)
```

```{r, echo=TRUE, include=FALSE}
# Find the average price of pumpkins per month
new_pumpkins %>% 
  group_by(month) %>% 
  summarize(price_mean = mean(price), na.rm = TRUE)
```


```{r, echo=TRUE, include=FALSE}
# Find the average price of pumpkins per month then plot a bar chart
pumpkins %>% 
  group_by(month) %>% 
  summarize(price_mean = mean(price), na.rm = TRUE) %>% 
  ggplot(aes(x = month, y = price_mean)) + geom_col()
```

```{r, include=FALSE}
# Specify a recipe
pumpkins_recipe <- recipe(price ~ ., data = new_pumpkins) %>% 
  step_integer(all_predictors(), zero_based = TRUE)


# Print out the recipe
pumpkins_recipe
```

```{r, include=FALSE}
# Prep the recipe
pumpkins_prep <- prep(pumpkins_recipe)

# Bake the recipe to extract a preprocessed new_pumpkins data
baked_pumpkins <- bake(pumpkins_prep, new_data = NULL)

# Print out the baked data set
baked_pumpkins %>% 
  slice_head(n = 10)
```

```{r, include=FALSE}
# Find the correlation between the package and the price
cor(baked_pumpkins$package, baked_pumpkins$price)
```

```{r,echo=TRUE, include=FALSE}
cor(baked_pumpkins$variety, baked_pumpkins$price)

cor(baked_pumpkins$day, baked_pumpkins$price)
```

```{r, include=FALSE}
# Load the corrplot package
library(corrplot)

# Obtain correlation matrix
corr_mat <- cor(baked_pumpkins %>% 

# Drop columns that are not really informative
select(-c(low_price, high_price)))

# Make a correlation plot between the variables
corrplot(corr_mat, method = "shade", shade.col = NA, tl.col = "black", tl.srt = 45, addCoef.col = "black", cl.pos = "n", order = "original")

```

```{r, include=FALSE}
set.seed(123)
# Split the data into training and test sets
pumpkins_split <- new_pumpkins %>% 
  initial_split(prop = 0.8)


# Extract training and test data
pumpkins_train <- training(pumpkins_split)
pumpkins_test <- testing(pumpkins_split)


# Create a recipe for preprocessing the data
lm_pumpkins_recipe <- recipe(price ~ package, data = pumpkins_train) %>% 
  step_integer(all_predictors(), zero_based = TRUE)


# Create a linear model specification
lm_spec <- linear_reg() %>% 
  set_engine("lm") %>% 
  set_mode("regression")
```

```{r, include=FALSE}
# Hold modelling components in a workflow
lm_wf <- workflow() %>% 
  add_recipe(lm_pumpkins_recipe) %>% 
  add_model(lm_spec)

# Print out the workflow
lm_wf
```

```{r, include=FALSE}
# Train the model
lm_wf_fit <- lm_wf %>% 
  fit(data = pumpkins_train)

# Print the model coefficients learned 
lm_wf_fit
```

```{r prediction_test, include=FALSE}
# Make predictions for the test set
predictions <- lm_wf_fit %>% 
  predict(new_data = pumpkins_test)


# Bind predictions to the test set
lm_results <- pumpkins_test %>% 
  select(c(package, price)) %>% 
  bind_cols(predictions)


# Print the first ten rows of the tibble
lm_results %>% 
  slice_head(n = 10)
```

```{r evaluate_lr, include=FALSE}
# Evaluate performance of linear regression
metrics(data = lm_results,
        truth = price,
        estimate = .pred)
```

```{r encode_package, include=FALSE}
# Encode package column
package_encode <- lm_pumpkins_recipe %>% 
  prep() %>% 
  bake(new_data = pumpkins_test) %>% 
  select(package)


# Bind encoded package column to the results
 plot_results <- lm_results %>%
 bind_cols(package_encode %>%
               rename(package_integer = package)) %>%
  relocate(package_integer, .after = package)


# Print new results data frame
plot_results %>%
  slice_head(n = 5)


# Make a scatter plot
plot_results %>%
  ggplot(mapping = aes(x = package_integer, y = price)) +
   geom_point(size = 1.6) +
   # Overlay a line of best fit
   geom_line(aes(y = .pred), color = "orange", size = 1.2) +
   xlab("package")
```

Today we will be continuing the pumpkin case study from last week. We will be using the data that you cleaned and split last time (pumpkins_train) and will be comparing our results today to those you have already obtained, so open and run your Lab 1 .Rmd as a first step so those objects are available in your Environment (unless you created an R Project last time, in which case, kudos to you!). 

Once you have done that, we'll start today's lab by specifying a recipe for a polynomial model.  First we specify a recipe that identifies our variables and data, converts package to a numerical form, and then add a polynomial effect with step_poly()

```{r}
# Specify a recipe
poly_pumpkins_recipe <-
  recipe(price ~ package, data = pumpkins_train) %>%
  step_integer(all_predictors(), zero_based = TRUE) %>% 
  step_poly(all_predictors(), degree = 2)
```

How did that work? Choose another value for degree if you need to. Later we will learn about model tuning that will let us do things like find the optimal value for degree.  For now, we'd like to have a flexible model, so find the highest value for degree that is consistent with our data.

Polynomial regression is still linear regression, so our model specification looks similar to before.

```{r, include=TRUE, echo=TRUE}
# Create a model specification called poly_spec
poly_spec <- linear_reg() %>% 
  set_engine("lm") %>% 
  set_mode("regression")
```
Question 1: Now take the recipe and model specification that just created and bundle them into a workflow called poly_df.

```{r, include=TRUE, echo=TRUE}
# Bundle recipe and model specification into a workflow
poly_df <- workflow() %>% 
  add_recipe(poly_pumpkins_recipe) %>% 
  add_model(poly_spec)
```

Question 2: fit a model to the pumpkins_train data using your workflow and assign it to poly_wf_fit
```{r, include=TRUE, echo=TRUE}
# Create a model using the workflow
poly_wf_fit <- poly_df %>% 
  fit(data = pumpkins_train)
```

```{r}
# Print learned model coefficients
poly_wf_fit
```

```{r, include=TRUE, echo=TRUE}
# Make price predictions on test data
poly_results <- poly_wf_fit %>% predict(new_data = pumpkins_test) %>% 
  bind_cols(pumpkins_test %>% select(c(package, price))) %>% 
  relocate(.pred, .after = last_col())

# Print the results
poly_results %>% slice_head(n = 5)
```

Now let's evaluate how the model performed on the test_set using yardstick::metrics().
```{r, include=TRUE, echo=TRUE}
metrics(data = poly_results, truth = price, estimate = .pred)

metrics(data = lm_results, truth = price, estimate = .pred)
```
Question 3: How do the performance metrics differ between the linear model from last week and the polynomial model we fit today?  Which model performs better on predicting the price of different packages of pumpkins?

**The polynomial regression is a better fit compared to the linear regression model and it also performs better. The rmse (root mean square error) describes the standard deviation of difference between predicted and observed values so the lower value in the polynomial regression means there are less errors and it is a better fit. For the rsq or coefficient of determination, values closer to 0 means the there is no relationship and values closer to 1 means there is a strong relationship and since the polynomial regression has a value closer to 1, it performs better at predicting the price.**

Let's visualize our model results.  First prep the results by binding the encoded package variable to them.
```{r, include=TRUE, echo=TRUE}
# Bind encoded package column to the results
poly_results <- poly_results %>% 
  bind_cols(package_encode %>% 
              rename(package_integer = package)) %>% 
  relocate(package_integer, .after = package)


# Print new results data frame
poly_results %>% 
  slice_head(n = 5)
```

OK, now let's take a look! 

Question 4: Create a scatter plot that takes the poly_results and plots package vs. price.  Then draw a line showing our model's predicted values (.pred). Hint: you'll need separate geoms for the data points and the prediction line.
```{r, include=TRUE, echo=TRUE}
# Make a scatter plot of the results 
poly_results %>% 
  ggplot(mapping = aes(x = package_integer, y = price)) +
   geom_point(size = 1.6, alpha = .2) +
   # Overlay a line of best fit
   geom_line(aes(y = .pred), color = "blue", size = 1.2, alpha = .8) +
   labs(x = "Package", y = "Price", title = "Package vs Price") + theme(plot.title = element_text(hjust = 0.5))

```

You can see that a curved line fits your data much better.

Question 5: Now make a smoother line by using geom_smooth instead of geom_line and passing it a polynomial formula like this:
geom_smooth(method = lm, formula = y ~ poly(x, degree = 3), color = "midnightblue", size = 1.2, se = FALSE)

```{r, include=TRUE, echo=TRUE}
# Make a smoother scatter plot 
poly_results %>% 
  ggplot(mapping = aes(x = package_integer, y = price)) +
   geom_point(size = 1.6, alpha = .2) +
   # Overlay a geom_smooth line
   xlab("package") + geom_smooth(method = lm, formula = y ~ poly(x, degree = 3), color = "midnightblue", size = 1.2, se = FALSE) + labs(x = "Package", y = "Price", title = "Package vs Price Smooth Plot") + theme(plot.title = element_text(hjust = 0.5))
```

OK, now it's your turn to go through the process one more time.
 
Additional assignment components :
6. Choose  a new predictor variable (anything not involving package type) in this dataset.

```{r}
# Names of all predictor variables
names(baked_pumpkins)
```

**My new predictor variable is variety.**

7. Determine its correlation with the outcome variable (price).  (Remember we calculated a correlation matrix last week)

```{r, include=TRUE, echo=TRUE}
# Calculate the correlation between pumpkin price and variety
cor(baked_pumpkins$variety, baked_pumpkins$price)

# Obtain correlation matrix and drop low_price and high
corr_matrix <- cor(baked_pumpkins %>% 
# Drop columns that are not really informative
  select(-c(low_price, high_price, month, day, date)))

# Make a correlation plot between the variables
corrplot(corr_matrix, method = "shade", shade.col = NA, tl.col = "black", tl.srt = 45, addCoef.col = "black", cl.pos = "n", order = "original")
```

8. Create and test a model for your new predictor:
  - Create a recipe
```{r, include=TRUE, echo=TRUE}
# Specify a recipe
pumpkins_recipe_variety <-
  recipe(price ~ variety, data = pumpkins_train) %>%
  step_integer(all_predictors(), zero_based = TRUE) 
#%>% step_poly(all_predictors(), degree = 3)
```
  
  - Build a model specification (linear or polynomial)
```{r, include=TRUE, echo=TRUE}
# Create a model specification
poly_spec_var <- linear_reg() %>% 
  set_engine("lm") %>% 
  set_mode("regression")
```
  
  - Bundle the recipe and model specification into a workflow
```{r, include=TRUE, echo=TRUE}
# Bundle recipe and model spec into a workflow
poly_wf_var <- workflow() %>% 
  add_recipe(pumpkins_recipe_variety) %>% 
  add_model(poly_spec_var)
```
  
  - Create a model by fitting the workflow
```{r, include=TRUE, echo=TRUE}
# Create a model
poly_wf_fit_var <- poly_df %>% 
  fit(data = pumpkins_train)
```

  - Evaluate model performance on the test data
```{r, include=TRUE, echo=TRUE}
# Make price predictions 
poly_results_var <- pumpkins_test %>% 
  select(c(variety, price)) %>% 
  bind_cols(predictions)

# Print the results
#poly_results_var %>% slice_head(n = 10)

# Evaluate model performance
metrics(data = poly_results_var, truth = price, estimate = .pred)
```
  
- Create a visualization of model performance
```{r, include=TRUE, echo=TRUE}
# Encode package column
encode_var <- pumpkins_recipe_variety %>%
  prep() %>%
  bake(new_data = pumpkins_test) %>%
  select(variety)

# Bind encoded package column to the results
 plot_results_var <- poly_results_var %>%
 bind_cols(encode_var %>%
               rename(variety_integer = variety)) %>%
  relocate(variety_integer, .after = variety)

# Print new results data frame
plot_results %>%
  slice_head(n = 5)

# Make a scatter plot
plot_results_var %>%
  ggplot(mapping = aes(x = variety_integer, y = price)) +
   geom_point(size = 1.6, alpha = .2) +
   # Overlay a line of best fit
   geom_line(aes(y = .pred), color = "orange", size = 1.2) +
   labs(x = "Variety", y = "Price", title = "Variety vs Price") + theme(plot.title = element_text(hjust = 0.5))


# Graph with geom_smooth
plot_results_var %>%
  ggplot(mapping = aes(x = variety_integer, y = price)) +
   geom_point(size = 1.6, alpha = .2) +
   # Overlay a geom_smooth line
    geom_smooth(method = lm, formula = y ~ poly(x, degree = 2), color = "orange", size = 1.2, se = FALSE) + labs(x = "Variety", y = "Price", title = "Variety vs Price Smooth Plot") + theme(plot.title = element_text(hjust = 0.5))
```

  
Lab 2 due 1/24 at 11:59 PM